GEE models

### 2016-07-06 doc/manuscript.Rmd
Made changes to manuscript as per tony's edits + prep for sub

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD

@@

Word count: 3602 / 4000

@@

<!-- Reference limit 41 / 40 -->

# Abstract
243 / 250 words

@@

sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).  See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))

@@

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], 
have been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

@@

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC. While the
final GEE model selected as best fitting differed between insulin sensitivity
and beta-cell function measures, the differences in QIC values were less than
10 between the models, suggesting similar fit. As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures (M8). The final GEE model was adjusted for time, sex, ethnicity,
baseline age, WC, ALT, and family history of diabetes. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every
SD increase in the predictor variable given the covariates are held constant.
Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
production, TAG may act as a mediator between NEFA and the outcomes. To
determine the role of TAG in the association between NEFA and the outcomes, TAG
was included in the GEE model as a sensitivity analysis. Lastly for the GEE
models, we tested for an interaction with sex, ethnicity, or time by the
predictor term for each outcome variable.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis, which was used to confirm the results
of the GEE modeling, but in a multivariate environment (all NEFA in a single
model). The PLS-DA results show a very poor discriminatory ability of the
NEFA composition in classifying participants into the correct beta-cell function
latent class. The percent of participants correctly predicted to belonging to
their group (i.e. those predicted in the low group and actually belonging to the
low group) was `r misclass.n.percent.bcf$Percent[9]` in the low group,
`r misclass.n.percent.bcf$Percent[5]` in the middle group, and 
`r misclass.n.percent.bcf$Percent[1]` in the high group (the distribution of the
groupings based on the components from PLS-DA are shown in `r cite_sf('plsda_groups')`). 
The loadings plot of the NEFA composition against the extracted PLS-DA
components is shown in `r cite_sf('plsda_loadings')` and is used to identify
fatty acids that contribute the most explained variance to the beta-cell
function groupings. However, because of the high rate of misclassification, no
meaningful predictive information can be extracted even if any NEFA does
contribute to the explained variance (e.g. 16:0). The PLS-DA analysis using
conversion to or maintenance of dysglycemia status over the 6 years as the
response variable showed similar poor discriminatory ability, though better than
the beta-cell function trajectory results. While 
`r misclass.n.percent.dysgly$Percent[1]`
of participants were correctly predicted to not have dysglycemia, only 
`r misclass.n.percent.dysgly$Percent[4]` were correctly predicted to have
dysglycemia (see `r cite_sf('plsda_groups_dysgly')` for the distribution of 
dysglycemia status based on the components from the PLS-DA and 
`r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
by extracted PLS-DA components). The results from the PLS-DA analysis are consistent
with what the associations seen in the GEE model.

@@

In a Canadian population of adults who are at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for the proportion of these fatty
acids and further confirmed by clustering analysis found no predictive ability of these
individual NEFA on beta-cell function or dysglycemia. These observations suggest
that the absolute size of the total NEFA fraction, rather than its specific
composition, likely influences the pathogenesis of diabetes, at least within a
population at-risk for diabetes.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified
WC as having a powerful attenuating effect on the associations between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, WC may be a strong causal link in previously reported
associations between NEFA and insulin sensitivity, given the role adipocytes
play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
in physiology between fasting and postprandial NEFA kinetics that we are not
able to investigate but that may explain our null findings for fasting NEFA and
insulin sensitivity. For instance, some experimental studies using clamp
protocols found that fasting NEFA was a weak predictor of insulin sensitivity
compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
metabolically active given postprandial activity. Third, the null findings seen
may be due to the high risk population examined in PROMISE, including having a 
high WC and BMI. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.

Few studies have examined the composition of NEFA on metabolic functioning. One 
recent, well-analyzed study used a variety of advanced fatty acid measurement 
and statistical techniques to explore the multivariate relationship between the 
NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a]. 
Specifically, the authors identified NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to
correlate with components of the MetS. Another similar study examining diabetes
found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers for identifying
healthy compared to diabetic individuals [@Liu2010a]. However, both studies were
limited by smaller sample sizes (approximately 100 subjects) and the
cross-sectional design.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures. However, we believe this
is a strength for our specific objective, as the chance of reverse causality is 
reduced given that fatty acid and glucose metabolism pathways are tightly 
integrated. This is also an observational cohort, and there may be some residual
confounding we have not considered or that could not be measured. Nonetheless,
potential covariates were empirically analyzed prior to inclusion into the
GEE models to best understand and minimizing potential
confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female of European-ancestry and as such our results may not be
generalizable to other populations. However, given these limitations, our study
also has several strengths, including the longitudinal design and the rigorous
statistical techniques and methods applied, which are specifically suited to
investigating temporal relationships and to handling the multivariate nature of
the data. Lastly, our cohort contains highly detailed and comprehensive
variable measurements at each collection visit, and has both concentration and
mol% data for the fatty acids.

@@

@@

```{r tabBasic}
table_basic(ds, caption = tab_basic)
```

@@

```{r figDist, fig.cap=fig_dist}
plot_nefa_distribution(ds)
```

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results, leg = 'none')
```

@@

The two plots generated from the PLS-DA analysis show the results from both
the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
the predictors (all 22 NEFA). The response result plots as seen in 
`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
showing an example of good discriminatory ability. When the predictors represent an
underlying pattern that may exist between groups, the plots will show greater
separation between the groups in addition to a low misclassification rate. The
predictor result plots as seen in `r cite_sf('plsda_loadings')` and
`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
contribute the most to the explained variance. Variables between the solid and 
dashed lines explain between 50-100% of the explained variance. Whether the 
variables have positive or negative values on either of the components axes 
determines how they contribute to each group. For instance, if those individuals
with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
has a positive value for component 1, this indicates that 16:0 is predictive of
having dysglycemia.

@@

```{r supTabQIC}
prep_qic <- gee_df %>%
    dplyr::select(
        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
        )

@@

```{r supTabDist}
table_distribution(ds, stab_dist)
```

```{r supFigHeatmap, fig.cap=sfig_corr}
plot_heatmap(ds)
```

```{r supFigGeeUnadj, fig.cap=sfig_gee}
plot_gee_results(gee_results_unadj)
```

```{r supTabGEE}
table_gee(gee_results, stab_gee)
table_gee(gee_results_unadj, stab_gee_unadj)
```

```{r supFigLCMM, fig.cap=sfig_lcmm}
plot_lcmm_results(lcmm_data)
```

```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results, 'Beta-cell function')
```

```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_results)
```

```{r supFigPlsGrpDysgly, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
```

```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_dysgly)
```

```{r supFigPlsExample, fig.cap=sfig_plsda_example}
plot_plsda_grouping(example_plsda_results(), 'Example groups')
```

>>>>

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,3,6), PhD

@@

Word count: 3620 / 4000

@@

<!-- Reference limit 39 / 40 -->

# Abstract
<!--243 / 250 words-->

@@

sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).  See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))

@@

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the metabolic impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], 
have been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as
eicosapentaenoic acid in the NEFA fraction would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

@@

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC. While the
final GEE model selected as best fitting differed between insulin sensitivity
and beta-cell function measures, the differences in QIC values were less than
10 between the models, suggesting similar fit. As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures (M8). The final GEE model was adjusted for time, sex, ethnicity,
baseline age, WC, ALT, and family history of diabetes. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every
standard deviation (SD) increase in the predictor variable given the covariates are held constant.
Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
production, TAG may act as a mediator between NEFA and the outcomes. To
determine the role of TAG in the association between NEFA and the outcomes, TAG
was included in the GEE model as a sensitivity analysis. Lastly for the GEE
models, we tested for an interaction with sex, ethnicity, or time by the
predictor term for each outcome variable.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis, which was used to confirm the results
of the GEE modeling, but in a multivariate environment (all NEFA in a single
model). The PLS-DA results show a very poor discriminatory ability of the
NEFA composition in classifying participants into the correct beta-cell function
latent class. The percent of participants correctly predicted to belonging to
their group (i.e. those predicted in the low group and actually belonging to the
low group) was `r misclass.n.percent.bcf$Percent[9]` in the low group,
`r misclass.n.percent.bcf$Percent[5]` in the middle group, and 
`r misclass.n.percent.bcf$Percent[1]` in the high group (the distribution of the
groupings based on the components from PLS-DA are shown in `r cite_sf('plsda_groups')`). 
The loadings plot of the NEFA composition against the extracted PLS-DA
components is shown in `r cite_sf('plsda_loadings')` and is used to identify
fatty acids that contribute the most explained variance to the beta-cell
function groupings. However, because of the high rate of misclassification, no
meaningful predictive information can be extracted even if any NEFA does
contribute to the explained variance (e.g. 16:0). The PLS-DA analysis using
conversion to or maintenance of dysglycemia status over the 6 years as the
response variable showed similar poor discriminatory ability, though better than
the beta-cell function trajectory results. While 
`r misclass.n.percent.dysgly$Percent[1]`
of participants were correctly predicted to not have dysglycemia, only 
`r misclass.n.percent.dysgly$Percent[4]` were correctly predicted to have
dysglycemia (see `r cite_sf('plsda_groups_dysgly')` for the distribution of 
dysglycemia status based on the components from the PLS-DA and 
`r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
by extracted PLS-DA components). The results from the PLS-DA analysis are consistent
with the associations seen in the GEE models.

@@

In a Canadian population of adults who are at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty
acids as a proportion, an observation that was further confirmed by multivariate
cluster techniques, which found no predictive ability of these
individual NEFA on beta-cell function or dysglycemia. These observations suggest
that the absolute size of the total NEFA fraction, rather than its specific
composition, likely influences the pathogenesis of diabetes, at least within a
population at-risk for diabetes.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified
WC as having a powerful attenuating effect on the associations between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, WC may be a strong causal link in previously reported
associations between NEFA and insulin sensitivity, given the role adipocytes
play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
in physiology between fasting and postprandial NEFA kinetics that we are not
able to investigate but that may explain our null findings for fasting NEFA and
insulin sensitivity. For instance, some experimental studies using clamp
protocols found that fasting NEFA was a weak predictor of insulin sensitivity
compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
metabolically active given postprandial activity. Third, the null findings seen
may be due to the high risk population examined in PROMISE, as the majority of subjects had a 
high WC and BMI. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.

Few studies have examined the composition of NEFA on metabolic functioning. One 
recent, well-analyzed study used a variety of advanced fatty acid measurement 
and statistical techniques to explore the multivariate relationship between the 
NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a]. 
Specifically, the authors identified NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to
correlate with components of the MetS. Another similar study examining diabetes
found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers for identifying
healthy compared to individuals with diabetes [@Liu2010a]. However, both studies were
limited by smaller sample sizes (approximately 100 subjects) and the
cross-sectional design.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures. However, we believe this
is a strength for our specific objective, as the chance of reverse causality is 
reduced given that fatty acid and glucose metabolism pathways are tightly 
integrated. This is also an observational cohort, and there may be some residual
confounding we have not considered or that could not be measured. Nonetheless,
potential covariates were empirically analyzed prior to inclusion into the
GEE models to best understand and minimize potential
confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and as such our results may not be
generalizable to other populations. However, given these limitations, our study
also has several strengths, including the longitudinal design and the rigorous
statistical techniques and methods applied, which are specifically suited to
investigating temporal relationships and to handling the multivariate nature of
the data. Lastly, our cohort contains highly detailed and comprehensive
variable measurements at each collection visit, and has both concentration and
mol% data for the fatty acids.

@@

```{r figure1, fig.cap=fig_dist}
plot_nefa_distribution(ds)
```

```{r figure2, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results, leg = 'none')
```

@@

```{r table1}
table_basic(ds, caption = tab_basic)
```

@@

@@

The two plots generated from the PLS-DA analysis show the results from both
the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
the predictors (all 22 NEFA). The response result plots as seen in 
`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
poor discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
showing a hypothetical example of good discriminatory ability. When the predictors represent an
underlying pattern that may exist between groups, the plots will show greater
separation between the groups in addition to a low misclassification rate. The
predictor result plots as seen in `r cite_sf('plsda_loadings')` and
`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
contribute the most to the explained variance. Variables between the solid and 
dashed lines explain between 50-100% of the explained variance. Whether the 
variables have positive or negative values on either of the components axes 
determines how they contribute to each group. For instance, if those individuals
with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
has a positive value for component 1, this indicates that 16:0 is predictive of
having dysglycemia.

@@

```{r onlineSupplementalTableS1}
prep_qic <- gee_df %>%
    dplyr::select(
        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
        )

@@

```{r onlineSupplementalTableS2}
table_distribution(ds, stab_dist)
```

```{r onlineSupplementalFigureS2, fig.cap=sfig_corr}
plot_heatmap(ds)
```

```{r onlineSupplementalFigureS3, fig.cap=sfig_gee}
plot_gee_results(gee_results_unadj)
```

```{r onlineSupplementalTableS3_4}
table_gee(gee_results, stab_gee)
table_gee(gee_results_unadj, stab_gee_unadj)
```

```{r onlineSupplementalFigureS4, fig.cap=sfig_lcmm}
plot_lcmm_results(lcmm_data)
```

```{r onlineSupplementalFigureS5, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results, 'Beta-cell function')
```

```{r onlineSupplementalFigureS6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_results)
```

```{r onlineSupplementalFigureS7, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
```

```{r onlineSupplementalFigureS8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_dysgly)
```

```{r onlineSupplementalFigureS9, fig.cap=sfig_plsda_example}
plot_plsda_grouping(example_plsda_results(), 'Example groups')
```
